Best Practices in Latent Class Analysis using the Free Open Source Software

### 2023-04-11 manuscript.tex
Final Proofread pt.2
Consistency checks, proofreading, updated table with sections

Third and final proofread to be pushed in by the end of workday today

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 12158}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1,2, Mauricio Garnier-Villarreal3, \& Daniel Anadria1},
  pdflang={en-EN},
  pdfkeywords={latent class analysis, mixture models, best practices, free open source software, tidySEM},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Best Practices in Latent Class Analysis using Free Open Source Software}
\author{Caspar J. van Lissa\textsuperscript{1,2}, Mauricio Garnier-Villarreal\textsuperscript{3}, \& Daniel Anadria\textsuperscript{1}}
\date{}

\shorttitle{BEST PRACTICES IN LATENT CLASS ANALYSIS}

@@

Correspondence concerning this article should be addressed to Caspar J. van Lissa, Padualaan 14, 3584CH Utrecht, The Netherlands. E-mail: \href{mailto:c.j.vanlissa@gmail.com}{\nolinkurl{c.j.vanlissa@gmail.com}}

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht\\\textsuperscript{3} Vrije Universiteit Amsterdam, Sociology}

\abstract{%
Latent class analysis (LCA) refers to a family of techniques for identifying groups
in data based on a parametric model.
Examples include mixture models, LCA with ordinal
indicators, and latent class growth analysis.
Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of LCA,
using free and open source software in R. To this end,
we introduce new functionality for estimating and reporting mixture models
using the \texttt{tidySEM} R-package.
}

@@

There are several compelling reasons to prefer free open source software instead.
FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
Furthermore, FOSS promotes collaboration within the scientific community,
as its users can contribute to the development, maintenance, and support of the software.
Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
researchers can review and verify the methods and algorithms used,
which helps to ensure the integrity and reliability of research findings.
Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.

@@

These limitations are addressed by the \texttt{tidySEM} package,
which provides a user-friendly interface for LCA in \texttt{OpenMx}.
The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting of LCA, as described in this paper.
The primary purpose of the \texttt{tidySEM} package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
The user interface is simple, as models can be specified using the widely used \texttt{lavaan} syntax (Rosseel, 2012).
The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
but the resulting models can still be fully customized making it a low threshold - high ceiling package.
For instance, LCA models constructed in \texttt{tidySEM} inherit the \texttt{MxModel} format from \texttt{OpenMx} which renders them compatible with various legacy functions from \texttt{OpenMx} and related LCA ecosystem .
Since \texttt{tidySEM} was developed with open science in mind, it facilitates fully-reproducible analyses and adheres to the \emph{FAIR (Findable, Accessible, Interoperable, and Reusable) software principles} (Lamprecht et al., 2020),
and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{\emph{OpenSSF Best Practices}} for FOSS software.

@@

The distribution is evidently bimodal, which bodes well for the intended
mixture model. In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that
the model estimates the mean shoe size of the two groups as equal to
9.22 and
7.25, which is close to the true
means of the two groups, namely 9.04
and 6.93. When tabulating estimated
group membership against observed (known) group membership, it can be
seen that women are classified with a high degree of accuracy, but men
are not:

@@

When the focus is on the model parameters in each group, LCA can be
thought of as similar to a multi-group structural equation model. The
main distinction is that group membership is not known a-priori, but estimated from the data.
Whereas in a multi-group model, the data are split by group and treated
as independent samples, in LCA, all cases contribute to the estimation
of all parameters in all groups. The relative contribution of each case
to the parameters of each group is determined by that case's \emph{posterior
probability} of belonging to that group.

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm that corrects for random measurement error by default, and is able to correct for systematic measurement error if it is included in the model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data given the class-specific model.
These posterior class probabilities can be used to weight follow-up analyses,
or to assign individuals to classes based on their highest class probability.
If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
This perspective on mixture modeling is sometimes referred to as ``model-based
clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016).
Many clustering algorithms apply some recursive splitting algorithm to the data. By
contrast ``model-based'' clustering refers to the fact that LCA estimates
cluster membership based on a parametric model.
Since a parametric model estimates only a fixed number of parameters,
it can be a relatively parsimonious solution compared to non-parametric techniques.

@@

The term LCA was initially conceived to refer to analyses with categorical,
usually binary indicators (Collins \& Lanza, 2009).
It was later generalized to refer to any technique that estimates a categorical latent variable based on a
parametric model of observed indicators, as it only refers
to the purpose of the analysis, and does not imply restrictions to the
model used, or the level of measurement of the indicators (Vermunt et al., 2004).
To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators},
and the sole \emph{LCA} expression as an umbrella term.
Given the abundance of redundant terms in use for closely related types of LCA models,
we here provide a rudimentary taxonomy.

A common type of LCA is the \emph{finite Gaussian mixture model}; a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be estimated
freely, or set to be constrained across classes (Collins \& Lanza, 2009).

The technique known as \emph{latent profile analysis (LPA)} is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for,
the residual covariances/correlations between indicators are
assumed to be zero.
In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes (Collins \& Lanza, 2009).
Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
a class of larger particles with a low correlation between length and width,
and a class of smaller particles with a high correlation between length and width.
The reason for this difference in correlations makes theoretical sense:
the large particles were heterogenous in shape,
whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).

A longitudinal extension of LCA is the \emph{Hidden Markov Model} (\emph{HMM}, also
called \emph{Latent Transition Analysis}), which estimates the change in class
membership over time under the Markov assumption. The Markov
assumption states that class membership at time \(t\) is only affected
by the class membership at time \(t-1\). The structural part of the model
is comprised of the initial states (proportion of the sample in each
class at the first time point), and transition probabilities
(probabilities of class membership switching across time). The transition
probabilities are estimated as multinomial logistic regressions between
class memberships (Vermunt, 2010).

@@

\hypertarget{testing-theory}{%
\subsubsection{Testing theory}\label{testing-theory}}

@@

\hypertarget{measurement-model}{%
\subsubsection{Measurement model}\label{measurement-model}}

@@

\hypertarget{dimension-reduction}{%
\subsubsection{Dimension reduction}\label{dimension-reduction}}

@@

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (Baughman, 2007; Magnus \& Garnier-Villarreal, 2022).
These models estimate one class to explain the ``true zeros'',
and another class with a different distribution to explain all other values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality (Alkema et al., 2022).
A mixture of two normal distributions with distinct variances described the observed distribution better.

@@

The best practices in estimation, as outlined in Table
\ref{tab:checkest-tab}, are rooted in existing recommendations
for estimating specific subtypes of LCA,
including latent
class growth analysis (Van De Schoot, Sijbrandij, Winter, Depaoli, \& Vermunt, 2017) and latent
class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
These guidelines were generalized to be relevant to all types of LCA,
and updated to current best practices, as explained below.

@@

\caption{\label{tab:checkest-tab}}
\centering
\begin{tabular}[t]{l|l}
\hline
\# & Item\\
\hline
1. & Examining Observed Data\\
\hline
2. & Handling Missing Data\\
\hline
3. & Alternative Model Specifications\\
\hline
4. & Software\\
\hline
5. & Algorithm\\
\hline
6. & Class Enumeration\\
\hline
7. & Model Fit Indices\\
\hline
8. & Classification Diagnostics\\
\hline
9. & Interpreting Class Solutions\\
\hline
10. & Label switching\\
\hline
\end{tabular}
\end{table}

@@

LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
Numeric variables should have an interval measurement level,
such that it is sensible to estimate theirmeans and (co)variances.
For the remaining three measurement levels,
some preprocessing is required.
The exact data types accepted by \texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}.
To correctly specify ordered variables
such as Likert scales,
use the function \texttt{OpenMx::mxFactor()}.
Binary variables,
such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using \texttt{mx\_dummies()}.

@@

Since SA is a global optimizer, it should be independent of starting values.
With this in mind, it is efficient to start estimation from a ``reasonable solution'', instead of a random starting point.
For example, if we assume that the different classes are likely to have different mean values on the indicators,
then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids (see also Biernacki, Celeux, \& Govaert, 2003).
By default, \texttt{tidySEM} derives starting values using K-means clustering (Likas, Vlassis, \& J. Verbeek, 2003),
computing the mixture model starting values by treating the K-means solution as multi-group model.
Next, SA is used to find the global optimum solution.
Finally, SA is followed up with a short run of the ML algorithm,
as ML inherently produces an asymptotic
covariance matrix for the parameters that can be used to compute
standard errors.
Note that these defaults can be manually overridden.

@@

Class enumeration refers to the process of determining the appropriate number of classes.
It is approached differently for exploratory versus confirmatory LCA.
In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes \(k\).
The number \(k\) may be chosen a-priori or dictated by the data.
It is sensible to limit \(k\) to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
Above a certain number of classes,
convergence problems typically arise,
or the proportion of cases assigned to the smallest classes may be low relative to the number of class-specific parameters.
In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
the proportion of cases assigned to the excessive classes converges to zero as the sample size increases (Rousseau \& Mengersen, 2011).
It is not known whether this also happens in frequentist LCA;
nonetheless, near-zero class memberships should be taken into account when determining the maximum \(k\).

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the \(-2LL\),
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The \emph{Akaike Information Criterion} (\emph{AIC}) is the original information criterion.
It is computed as \(-2LL+2p\), where \(p\) is the number of parameters.
The \emph{Bayesian Information Criterion} (\emph{BIC}) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
such that complex models are penalized more in smaller samples.
Finally,
the so-called \emph{sample size adjusted BIC} (\emph{saBIC}) implements a different ad-hoc penalty for sample size,
\(\ln(\frac{n + 2}{24})\). Although simulation studies have showen that the saBIC performs well for class enumeration (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013),
a limitation is that its penalty is not based on theory. BIC also performs well and is theoretically substantiated.
As such, BIC may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
In \texttt{tidySEM}, all three ICs are available as a part of the output of \texttt{table\_fit()}.

@@

The first author's anecdotal experience as a statistical co-author on LCA studies suggests that both aforementioned tests tend to be liberal,
returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
Other authors have reported similar experiences (Sinha, Calfee, \& Delucchi, 2021), which matches the differences between LRT and ICs from information theory as LRT tests and compare the models based on in sample predictive accuracy, and ICs evaluates and compare the models out-of-sample predictive accuracy which helps to control for overfitting (McElreath, 2020).
This suggests that, like existing ICs, LR tests might not adequately balance fit and complexity in LCA.
Better solutions for class enumeration are thus an important target for future research.

@@

Second, most likely class membership based on the highest posterior class probability, \(F_M\), is obtained by running \texttt{class\_prob(res,\ "sum.mostlikely")}.
This is the frequency table of the variable \(M\).
Unlike \(F_P\), \(F_M\) ignores classification error.
If every observation is classified with perfect accuracy,
\(F_M\) and \(F_P\) are identical.
Part of the output of \texttt{table\_fit()} for LCA analyses
are the minimum and maximum percentage of the sample assigned to a particular class, \texttt{n\_min} and \texttt{n\_max}, derived from \(F_M\).
This is reported by default because the smallest class proportion, \texttt{n\_min}, is of particular interest.
When the size of the smallest class is not much larger than the number of parameters per class,
proceed with caution:
The model might not be locally identified in that class (Depaoli, 2013).
Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
This is not particularly meaningful,
however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
Thus, a better rule of thumb might be to ascertain that the smallest class has many more observations than the average number of parameters per class.
Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
If the smallest class does not comprise at least a few observations per parameter, or is redundant with another larger class, a simpler solution is preferable.

To assess classification accuracy, we turn to the third diagnostic table:
the average posterior probabilities by most likely class membership, \(P_{p_{(M = m)}}\)
which can be obtained using \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}.
For participants assigned to one particular class (rows) according to \(M\),
this table gives the mean probability of being assigned to any class (columns).
This table is obtained by calculating the column means of \(P\) for the
\(k\) subsets of observations with most likely class of \(M = 1 \ldots k\).
The diagonal represents the average predictive probability of the subjects related
to their most likely class, indicating the certainty of class assignment.
Some guidelines have being suggested for this diagonal,
such as its values having to be higher than 0.7 (Masyn, 2013).
We, however, advise caution as these guidelines cannot be generalized to all situations,
and should not be used as strict cutoffs communication the classification quality.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 12158}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Best Practices in Latent Class Analysis using the Free Open Source Software},
  pdfauthor={Caspar J. van Lissa1, Mauricio Garnier-Villarreal2, \& Daniel Anadria3},
  pdflang={en-EN},
  pdfkeywords={latent class analysis, mixture models, best practices, free open source software, tidySEM},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Best Practices in Latent Class Analysis using the Free Open Source Software}
\author{Caspar J. van Lissa\textsuperscript{1}, Mauricio Garnier-Villarreal\textsuperscript{2}, \& Daniel Anadria\textsuperscript{3}}
\date{}

\shorttitle{BEST PRACTICES LATENT CLASS ANALYSIS}

@@

The authors made the following contributions. Caspar J. van Lissa: Conceptualization, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Mauricio Garnier-Villarreal: Writing -- review \& editing; Daniel Anadria: Writing -- original draft, Writing -- review \& editing.

Correspondence concerning this article should be addressed to Caspar J. van Lissa, Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands. E-mail: \href{mailto:c.j.vanlissa@tilburguniversity.edu}{\nolinkurl{c.j.vanlissa@tilburguniversity.edu}}

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, Methodology \& Statistics\\\textsuperscript{2} Vrije Universiteit Amsterdam, Sociology\\\textsuperscript{3} Utrecht University, Methodology \& Statistics}

\abstract{%
Latent class analysis (LCA) refers to a family of techniques for identifying groups
in data based on a parametric model.
Examples include mixture models, LCA with ordinal
indicators, and latent class growth analysis.
Despite the popularity of this technique, there is limited
guidance with respect to best practices in estimating and reporting mixture
models. Moreover, although user-friendly interfaces for advanced mixture
modeling have long been available in commercial software packages, open
source alternatives have remained somewhat inaccessible. This tutorial
describes best practices for the estimation and reporting of LCA,
using the free and open source software in R. To this end,
we introduce new functionality for estimating and reporting mixture models
using the \texttt{tidySEM} R-package.
}

@@

There are several compelling reasons to prefer free open source software instead.
FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
Furthermore, FOSS promotes collaboration within the scientific community,
as its users can contribute to the development, maintenance, and support of the software.
Finally, FOSS promotes transparency and reproducibility: Because the source code for FOSS is accessible,
researchers can review and verify the methods and algorithms used,
which helps to ensure the integrity and reliability of research findings.
Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.

@@

These limitations are addressed by the \texttt{tidySEM} package,
which provides a user-friendly interface for LCA in \texttt{OpenMx}.
The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting of LCA, as described in this paper.
The primary purpose of the \texttt{tidySEM} package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
The user interface is simple, as models can be specified using the widely used \texttt{lavaan} syntax (Rosseel, 2012).
The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
but the resulting models can still be fully customized making it a low threshold high ceiling package.
For instance, LCA models constructed in \texttt{tidySEM} inherit the \texttt{MxModel} format from \texttt{OpenMx} which renders them compatible with various legacy functions from \texttt{OpenMx} and related LCA ecosystem .
Since \texttt{tidySEM} was developed with open science in mind, it facilitates fully-reproducible analyses and adheres to the \emph{FAIR (Findable, Accessible, Interoperable, and Reusable) software principles} (Lamprecht et al., 2020),
and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{\emph{OpenSSF Best Practices}} for FOSS software.

@@

The distribution is evidently bimodal, which bodes well for the intended
mixture model. In this case, the number of classes is known a priori.
When estimating a two-class mixture model, the detective observes that
the model estimates the mean shoe size of the two groups as equal to
9.22 and
7.25, which is close to the true
means of the two groups, namely 9.04
and 6.93. When tabulating estimated
group membership against observed (known) group membership, it can be
seen that women are classified with a high degree of accuracy, but men
are not:

@@

When the focus is on the model parameters in each group, LCA can be
thought of as similar to a multi-group structural equation model. The
main distinction is that group membership is not known a priori, but estimated from the data.
Whereas in a multi-group model, the data are split by group and treated
as independent samples, in LCA, all cases contribute to the estimation
of all parameters in all groups. The relative contribution of each case
to the parameters of each group is determined by that case's \emph{posterior
probability} of belonging to that group.

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm that corrects for random measurement error by default, and is able to correct for systematic measurement error if it is included in the model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data given the class-specific model.
These posterior class probabilities can be used to weight follow-up analyses,
or to assign individuals to classes based on their highest class probability.
The latter approach may be acceptable when the classes are clearly distinct and the measurement error is low.
This perspective on mixture modeling is sometimes referred to as \emph{model-based
clustering} (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016).
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast, model-based clustering refers to the fact that LCA estimates
cluster membership based on a parametric model.
Since parametric models estimate only a fixed number of parameters,
they can be a relatively parsimonious solution compared to non-parametric techniques.

@@

Initially, the term LCA was conceived to refer to analyses with categorical,
usually binary indicators (Collins \& Lanza, 2009).
Later, it became generalized to refer to any technique that estimates a categorical latent variable based on a
parametric model of observed indicators, as the term only refers
to the purpose of the analysis, and does not imply restrictions to the
model used, or the level of measurement of the indicators (Vermunt et al., 2004).
To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators},
and use \emph{LCA} as an umbrella term for the family techniques.
Given the abundance of redundant terms in use for closely related types of LCA models,
we here provide a rudimentary taxonomy.

A common type of LCA is the \emph{finite Gaussian mixture model}, a
univariate analysis where the observed distribution of a single variable
is assumed to result from a mixture of a known number of normal
(Gaussian) distributions. The parameters of a finite Gaussian mixture
model are the means and variances of these underlying normal
distributions. The analysis of shoe sizes presented earlier is a
canonical example of this type of analysis. In the multivariate case,
with more than one indicator variable, the parameters of a mixture model
are the means, variances, and covariances between the indicators (which
can be standardized to obtain correlations). These parameters can be estimated
freely, or set to be constrained across classes (Collins \& Lanza, 2009).

The technique known as \emph{latent profile analysis (LPA)} is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for,
the residual covariances/correlations between indicators are
assumed to be zero.
In some cases such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes (Collins \& Lanza, 2009).
Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
a class of larger particles with a low correlation between length and width,
and a class of smaller particles with a high correlation between length and width.
The reason for this difference in correlations makes theoretical sense:
the large particles were heterogenous in shape,
whereas the smaller particles had been polished to a more uniform, rounded shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).

A longitudinal extension of LCA is the \emph{Hidden Markov Model} (\emph{HMM}, also
called \emph{Latent Transition Analysis}), which estimates the change in class
membership over time under the \emph{Markov assumption}. The Markov
assumption states that class membership at time \(t\) is only affected
by the class membership at time \(t-1\). The structural part of the model
is comprised of the initial states (proportion of the sample in each
class at the first time point), and transition probabilities
(probabilities of class membership switching across time). The transition
probabilities are estimated as multinomial logistic regressions between
class memberships (Vermunt, 2010).

@@

\hypertarget{testing-theory}{%
\subsubsection{Testing Theory}\label{testing-theory}}

@@

\hypertarget{measurement-model}{%
\subsubsection{Measurement Model}\label{measurement-model}}

@@

\hypertarget{dimension-reduction}{%
\subsubsection{Dimension Reduction}\label{dimension-reduction}}

@@

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (Baughman, 2007).
These models estimate one class to explain the ``true zeros'',
and another class with a different distribution to explain all other values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
One example of mixture models for violations of normality
is in modeling responses to the PH-9, a questionnaire of depressive symptoms.
Researchers observed two distinct groups: One with no symptoms on most indicators (zero inflation),
and one with non-zero symptoms (Magnus \& Garnier-Villarreal, 2022).
As this response pattern violates the assumption of multivariate normality,
the authors used a mixture model to capture the ``presence'' and ``frequency'' of symptoms as separate classes.

@@

The best practices in estimation, as outlined in Table
\ref{tab:checkest-tab}, are rooted in existing recommendations
for estimating specific subtypes of LCA,
including latent
class growth analysis (Van De Schoot, Sijbrandij, Winter, Depaoli, \& Vermunt, 2017) and latent
class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
These guidelines were generalized to be relevant to all types of LCA,
and updated to current best practices showcasing \texttt{tidySEM} syntax.

@@

\caption{\label{tab:checkest-tab}}
\centering
\begin{tabular}[t]{l|l}
\hline
\# & Item\\
\hline
1. & Examining Observed Data\\
\hline
2. & Data Preprocessing\\
\hline
3. & Handling Missing Data\\
\hline
4. & Model Specification\\
\hline
5. & Estimation and Convergence\\
\hline
6. & Class Enumeration\\
\hline
7. & Model Fit Indices\\
\hline
8. & Nested Model Tests\\
\hline
9. & Classification Diagnostics\\
\hline
10. & Labelling Classes\\
\hline
11. & Follow-up Analyses\\
\hline
\end{tabular}
\end{table}

@@

LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
Numeric variables should have an interval measurement level,
such that it is sensible to estimate their means and (co)variances.
For the remaining three measurement levels,
some preprocessing is required.
The exact data types accepted by \texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}.
To correctly specify ordered variables
such as Likert scales,
use the function \texttt{OpenMx::mxFactor()}.
Binary variables,
such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using \texttt{mx\_dummies()}.

@@

Since SA is a global optimizer, it should be independent of starting values.
With this in mind, it is efficient to start estimation from a ``reasonable solution'', instead of a random starting point.
For example, if we assume that the different classes are likely to have different mean values on the indicators,
then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids (see also Biernacki, Celeux, \& Govaert, 2003).
By default, \texttt{tidySEM} derives starting values using K-means clustering (Hartigan \& Wong, 1979),
computing the mixture model starting values by treating the K-means solution as multi-group model.
Next, SA is used to find the global optimum solution.
Finally, SA is followed up with a short run of the ML algorithm,
as ML inherently produces an asymptotic
covariance matrix for the parameters that can be used to compute
standard errors.
Note that these defaults can be manually overridden.

@@

Class enumeration refers to the process of determining the appropriate number of classes.
It is approached differently for exploratory versus confirmatory LCA.
In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes \(k\).
The number \(k\) may be chosen a priori or dictated by the data.
It is sensible to limit \(k\) to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
Above a certain number of classes,
convergence problems typically arise,
or the proportion of cases assigned to the smallest classes may be low relative to the number of class-specific parameters.
In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
the proportion of cases assigned to the excessive classes converges to zero as the sample size increases (Rousseau \& Mengersen, 2011).
It is not known whether this also happens in frequentist LCA;
nonetheless, near-zero class memberships should be taken into account when determining the maximum \(k\).

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the \(-2LL\),
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The \emph{Akaike Information Criterion} (\emph{AIC}) is the original information criterion.
It is computed as \(-2LL+2p\), where \(p\) is the number of parameters.
The \emph{Bayesian Information Criterion} (\emph{BIC}) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
such that complex models are penalized more in smaller samples.
Finally,
the so-called \emph{sample size adjusted BIC} (\emph{saBIC}), \(\ln(\frac{n + 2}{24})\),
implements a penalty for sample size that is based on the Minimum Description Length, a concept from data compression that refers to the smallest number of pieces of information necessary to reproduce the data (Rissanen, 1983; Sclove, 1987).
Simulation studies have shown that the saBIC performs well for enumeration of homogenous clusters in small samples,
but tends to select too many classes when the number of the number of classes and the sample size are large (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013).
The BIC has been found to perform consistently well across simulation studies,
and is among the most widely known information criteria.
As such, BIC may be a sensible default criterion for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
In \texttt{tidySEM}, all three ICs are available as a part of the output of \texttt{table\_fit()}.

@@

There are known limitations to both aforementioned tests;
on the one hand, the LRT is known to contain mathematical inaccuracies (Jeffries, 2003), yet still shows reasonable performance in simulation studies.
On the other hand, the BLRT is particularly sensitive to model misspecification and is thus not appropriate for exploratory applications (Nylund et al., 2007).
The first author's anecdotal experience as a statistical co-author on LCA studies suggests that both aforementioned tests tend to be liberal,
returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
Other authors have reported similar experiences (Sinha, Calfee, \& Delucchi, 2021), which matches the differences between LRT and ICs from information theory as LRT tests and compare the models based on in sample predictive accuracy, and ICs evaluates and compare the models out-of-sample predictive accuracy which helps to control for overfitting (McElreath, 2020).
This suggests that, like existing ICs, LR tests might not adequately balance fit and complexity in LCA.
Better solutions for class enumeration are thus an important target for future research.

@@

Second, most likely class membership based on the highest posterior class probability, \(F_M\), is obtained by running \texttt{class\_prob(res,\ "sum.mostlikely")}.
This is the frequency table of the variable \(M\).
Unlike \(F_P\), \(F_M\) ignores classification error.
If every observation is classified with perfect accuracy,
\(F_M\) and \(F_P\) are identical.
Part of the output of \texttt{table\_fit()} for LCA analyses
are the minimum and maximum percentage of the sample assigned to a particular class, \texttt{n\_min} and \texttt{n\_max}, derived from \(F_M\).
This is reported by default because the smallest class proportion, \texttt{n\_min}, is of particular interest.
When the size of the smallest class is not much larger than the number of parameters per class,
proceed with caution:
The model might not be locally identified in that class (Depaoli, 2013).
Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
A major concern with these rules of thumb is that it ignores the absolute sample size within each class.
Small classes may comprise too few observations to accurately estimate class parameters.
Thus, a better rule of thumb might be to ascertain that the smallest class is sufficiently large;
both in terms of absolute sample size and in terms of the number of observations per estimated parameter.
Well-established rules of thumb suggest ratios of 10 or 20 observations per parameter (Jackson, 2003; Kline, 2016).
Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
If the smallest class has too few observations per parameter, or its substantive interpretation is redundant with another class,
a simpler solution with fewer classes or a simpler class-specific model is preferable.
